---
title: "COVID Project"
author: "Team Delta"
date: '2020-10-20'
output:
  github_document:
    toc: yes
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
---

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, include=FALSE}
library(tidyverse)
```

```{r read-data}
filename_data <- "../data/combined_data.csv"
df_data <- read_csv(filename_data)
df_data
```

# EAMON EDA

**Goal**: Try comparing numbers of cases / deaths at day of measure implemented for all states.

```{r date_vs_covid}
df_data %>%
  drop_na(num_measures_perday) %>%
  ggplot(aes(date, total_cases)) +
  geom_point(aes(size = num_measures_perday), alpha=0.5)

df_data %>%
  drop_na(num_measures_perday) %>%
  ggplot(aes(date, total_deaths)) +
  geom_point(aes(size = num_measures_perday), alpha=0.5)
```

**Observations**:

- Generally clustered around low amounts of low cases / deaths
  - Follows the observation that most measures were early on in the pandemic
- Some outliers
  - A couple clear exponential (?) lines coming out of the main cluster.
    - Expect each line to be a single state, would be interesting if it is. Would also show a correlation between cases and measures, although this could likely be attributed to cases increasing over time, and these states specifically consistently giving out measures. (Not like measures are increasing in count of anything)
  - Also a single flat line coming out of the cluster, which is likely a state not hit hard but still releasing measures? Potentially interesting.
- Seems the most dense mid / late March to April, might be worth looking into straight distribution over time?
- Deaths seem a bit more clustered than cases - might be be indicative of deaths being a greater driver of decisions?

**Goal**: Sanity check - plot normalized cases and deaths

```{r date_vs_covid_norm}
df_data %>%
  drop_na(num_measures_perday) %>%
  ggplot(aes(date, cases_per100k)) +
  geom_point(aes(size = num_measures_perday), alpha=0.5)

df_data %>%
  drop_na(num_measures_perday) %>%
  ggplot(aes(date, deaths_per100k)) +
  geom_point(aes(size = num_measures_perday), alpha=0.5)
```

**Observations**:

- Death is no longer "more clustered" than cases.
- No states with no cases or deaths past a certain point.
- See a clear bottom line in terms of cases / deaths
  - However, no real relation between this line and the measures, so likely just showing covid spreading
  
**Goal**: Validate whether or not each line is a state.

```{r date_vs_covid_vs_state, fig.width=12}
df_data %>%
  drop_na(num_measures_perday) %>%
  ggplot(aes(date, cases_per100k)) +
  geom_point(aes(size = num_measures_perday, color=state))

df_data %>%
  drop_na(num_measures_perday) %>%
  ggplot(aes(date, deaths_per100k)) +
  geom_point(aes(size = num_measures_perday, color=state))
```

**Observations**:

- Yes, each line corresponds to a state.

**Goal**: Plot distribution of measures over time

```{r measures_distribution}
df_data %>%
  drop_na(num_measures_perday) %>%
  ggplot(aes(date, num_measures_perday)) +
  geom_col()
```

**Observations**:

- Shows a clear cluster of measures around mid March, distribution seems about normal with a tail to the right.

**Goal**: Check if interventions are made in response to the recent number of cases

```{r avg_cases_vs_meaures}
slideMean <- function(x, windowsize=3, slide=1){
 idx1 <- seq(1,length(x),by=slide);
 idx1 + windowsize -> idx2;
 idx2[idx2 > (length(x) + 1)] <- length(x) + 1;
 c(0, cumsum(x)) -> cx;
 ((cx[idx2] - cx[idx1]) / windowsize) -> cum_avg
 c(rep(0, windowsize - 1), cum_avg[1:(length(cum_avg) - windowsize + 1)]) -> trim_avg
 return(trim_avg);
}

window_size = 14
df_data %>%
  group_by(state) %>%
  mutate(avg_cases = slideMean(cases_per100k, window_size)) %>%
  slice(window_size:n()) %>%
  ungroup() %>%
  drop_na(num_measures_perday) %>%
  ggplot(aes(avg_cases, num_measures_perday)) +
  geom_point()
```

**Observations**:

- Possibly a weak negative relationship between avg cases number of measures? So, the more average cases there are, the less measures there are.
  - Seems likely in the case that we have more average cases later in the pandemic, and more average cases earlier in the pandemic
  - Seems like somehow accounting for the increase in cases over the course of the pandemic is important - maybe normalize by the total so far?
  
**Goal**: Plot average cases over time and measures over time to get a sense of how they match up.

```{r avg_cases_vs_measures_vs_time}
# slideMean(head(df_data, 20) %>% pull(cases_per100k), windowsize=14)
# head(df_data) %>% pull(cases_per100k)

window_size = 7
df_data %>%
  filter(date < "2020-7-1") %>%
  group_by(state) %>%
  mutate(avg_cases = slideMean(cases_per100k, windowsize=window_size)) %>%
  ungroup() %>%
  ggplot(aes(date, avg_cases)) +
  geom_point() +
  geom_col(aes(date, num_measures_perday * 5)) +
  scale_y_continuous(sec.axis = sec_axis(~./25, name="Num measures"))

df_data %>%
  filter(date < "2020-7-1") %>%
  group_by(state) %>%
  mutate(avg_cases = slideMean(cases_per100k, windowsize=window_size)) %>%
  ungroup() %>%
  ggplot(aes(date, avg_cases)) +
  geom_smooth() +
  geom_col(aes(date, num_measures_perday * 5)) +
  scale_y_continuous(sec.axis = sec_axis(~./25, name="Num measures"))

df_data %>%
  filter(date < "2020-7-1") %>%
  group_by(state) %>%
  mutate(avg_cases = slideMean(cases_per100k, windowsize=window_size)) %>%
  ungroup() %>%
  group_by(date) %>%
  mutate(avg_avg_cases = mean(avg_cases)) %>%
  ggplot(aes(date, avg_avg_cases)) +
  geom_col(aes(date, num_measures_perday * 5), color = "blue", fill = "blue") +
  geom_smooth(color = "black") +
  scale_y_continuous(
    name = "Average of Rolling Average for each State",
    sec.axis = sec_axis(~./25, name="Num measures")
  ) +
  scale_x_date(name = "Date") +
  theme(
      axis.title.y = element_text(color = "black"),
      axis.title.y.right = element_text(color = "blue")
  )
```

**Observations**:

- Seems like a majority of the measures were taken right before the cases spiked, which would mean most of the cases would have very low average cases.
- Generally bad new for this direction. If most of the measures are made right before cases spike, then theres really not much info to look at.

**Goal**: Plot just cases over time vs measures and see how they match up

```{r cases_vs_measures_vs_time, fig.width = 15}
df_data %>%
  filter(date < "2020-7-1") %>%
  group_by(date) %>%
  summarise(
    avg_total_cases = mean(cases_per100k),
    total_measures = sum(num_measures_perday, na.rm = TRUE) * 5
  ) %>%
  ungroup() %>%
  ggplot(aes(date, avg_total_cases)) +
  geom_col(aes(date, total_measures), color = "blue", fill = "blue") +
  geom_point(color = "black") +
  scale_y_continuous(
    name = "Average Total Cases",
    sec.axis = sec_axis(~./5, name="Num Measures")
  ) +
  scale_x_date(name = "Date") +
  theme(
      axis.title.y = element_text(color = "black"),
      axis.title.y.right = element_text(color = "blue")
  )

df_total_measures <-
  df_data %>%
  filter(date < "2020-7-1") %>%
  group_by(date) %>%
  summarise(
    avg_total_cases = mean(cases_per100k),
    total_measures = sum(num_measures_perday, na.rm = TRUE) * 10
  ) %>%
  ungroup()

df_data %>%
  filter(date < "2020-7-1") %>%
  group_by(state, date) %>%
  ungroup() %>%
  ggplot(aes(date, cases_per100k, color = state)) +
  geom_col(
    data = df_total_measures,
    aes(y = total_measures), color = "black", fill = "black") +
  geom_point() +
  scale_y_continuous(
    name = "Average Cases",
    sec.axis = sec_axis(~./10, name="Num Measures")
  ) +
  scale_x_date(name = "Date") +
  theme(
      axis.title.y = element_text(color = "black"),
      axis.title.y.right = element_text(color = "black"),
      legend.position="none"
  )
```

**Observations**:

- Issue persists even when not looking at averages. Not good, means there is very little pre-measure data to pull trends from.